Gametic Selection, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2016-09-25 LaTeX/Turnover.tex
comparing to Kirk

%life-cycle
The life-cycle begins with competition between haploid gametes/gametophytes from each sex, where selection depends on the sex of the diploid they came from and their allele at the \textbf{A} locus.
Gametes/gametophytes from males then mate randomly with gametes/gametophytes from females.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (or the \textbf{M} genotype of their mother).
Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
This is followed by meiosis with recombination. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$. %is chi really the prob of rec in our recursion equations?
Any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$.
We track the frequency of haploid genotypes produced by each sex from one generation to the next (recursion equations in Sup.\ Mat.).

@@

\subsection*{Resident equilibrium and stability}
With allele $M$ fixed, we follow the dynamics of the frequency of $A$ in gametes/gametophytes from the homogametic sex (e.g., in eggs from an XX female, $p^f_X$) and the frequencies of $A$ in the two types of gametes/gametophytes from the heterogametic sex (e.g., in sperm from an XY male that are X-bearing, $p^m_X$, and Y-bearing, $p^m_Y$).
Assuming selection is weak relative to recombination, the differences in the frequencies of $A$ in each type of gamete/gametophyte are small, and we can solve for the mean frequency of $A$ across all types ($p_A$) as well as the difference in the frequencies of $A$ between two of the three types, to first order in selection.
Linear stability analysis can then be used to determine the stability of this equilibrium.
With no haploid selection our results reduce to those of \cite{vanDoorn:2010hu} when $k=0$ (neo-$Y$ invading a $ZW$ system) or when $k=1$ (neo-$W$ invading an $XY$ system).

\subsection*{Sex chromosome turnover}
The spread of a rare mutant $m$ at the \textbf{M} locus in such a population is determined by the leading eigenvalue, $\lambda$, of the system described by the equations for the next generation frequency of haploid genotypes with the mutation. %, evaluated at the above equilibrium.
Below we present the results for invasion into an $XY$ system assuming no competition among gametes/gametophytes from females and linear arrangement \textbf{X}\textbf{A}\textbf{M}. %(SAY SOMETHING HERE ABOUT OTHER ARRANGEMENTS?). %equal dominance coefficients in the two sexes,
Equivalent results for invasions into a $ZW$ system can be derived by consistently switching the roles of males and females \citep{vanDoorn:2010hu}.

@@

Otherwise, the new sex-determining allele increases in frequency on one \textbf{A} background and declines on the other and invasion requires
\begin{equation}\label{eq:neoYR}
R \left[ \frac{p^f_X w^m_a}{g_A^*} + \frac{(1 - p^f_X) w^m_A}{g_a^*} \right] w^m_{Aa} < \nu^m,
\end{equation}
for the neo-$Y$, and 
\begin{equation}\label{eq:neoWR}
R \left[  \frac{(1 - \bar{p}^m) w^m_{a}}{g_A^*} + \frac{\bar{p}^m w^m_{A}}{g_a^*} \right] w^f_{Aa} < \nu^f
\end{equation}
for the neo-$W$.
Here $\bar{p}^m = (p^m_X+p^m_Y)/2$ is the average frequency of $A$ in males, $w^i_{j}$ is the relative viability fitness of sex $i\in\{m,f\}$ depending on their haploid or diploid genotype at the \textbf{A} locus (with $j\in\{A,a,AA,Aa,aa\}$) and
$\nu^i = 
  p^f_X p^m_{\textbf{X}^i} w^m_{a} w^i_{AA} +
  p^f_X (1-p^m_{\textbf{X}^i}) w^m_{a} w^i_{Aa} +
  (1-p^f_X) p^m_{\textbf{X}^i} w^m_{a} w^i_{Aa} +
  (1-p^f_X) (1-p^m_{\textbf{X}^i}) w^m_a w^i_{aa}$,
with $\textbf{X}^m = Y$ and $\textbf{X}^f = X$, is the mean relative fitness of resident individuals of sex $i$.
%and $g_j^*$ is the growth rate of $X^m$ and $Ym$ haplotypes when on an $A$ ($j=A$)  or $a$ background ($j=a$).

The new sex-determining allele is therefore expected to invade for any recombination rate, $R$, when the net flow of double recombinants is from the less fit to the more fit \textbf{A} background (making the term inside the square brackets negative).
%Assuming $g_A^* > 0$ and $g_a^* < 0$ without loss of generality, this is $-(1 - p^f_X) w_{m,A} g_a^* < p^f_X w_{m,a} g_A^*$.
When the net flow of double recombinants is from the more fit to the less fit haplotype, the new sex-determining allele can still invade when the rate of recombination between it and the selected locus is small enough.

%weak selection
Assuming weak selection, we can solve for the invasion fitness of the new sex -determining alleles explicitly, giving
\begin{equation}
\lambda_{m,XY} \approx 1 + V_A S_A C_m,
\end{equation}
where $V_A = p_A(1-p_A)$ is the variance at the \textbf{A} locus, $S_A = s^f (t^m)^2 / \left[ r R \left( s^f + s^m \right)^2 \right]$ (\textcolor{red}{INTERPRETATION? LOSE THE R?}), $s^f$ and $s^m$ are the respective selection coefficients for $A$ in diploid females and males, $t^m$ is the selection coefficient for $A$ in gametes/gametophytes from males, and we've assumed equal dominance coefficients in the two sexes.

For the neo-$Y$ we have $C_Y = (r - R)s^f $.
The neo-$Y$ can therefore invade whenever it is in tighter linkage with the selected locus than the ancestral sex-determining locus, $r>R$, provided locus \textbf{A} is polymorphic ($V_A>0$) and there is selection among both females ($s^f\neq0$) and gametes/gametophytes from males ($t^m\neq0$).

For the neo-$W$ we have $C_W = [(2r(1-R)-R) s^f + (1-2r)R s^m]/2$.
In this case, even when the novel sex-determining locus is in looser linkage with the selected locus than is the ancestral sex-determining locus, $r<R$, a neo-$W$ can invade.
For example, with $R=1/2$ the neo-$W$ invades if there is any linkage between the ancestral sex-determining and selected loci ($r<1/2$), there is selection among gametes/gametophytes in males ($t^m\neq0$), and there is selection for or against $A$ in both males and females ($s^m s^f >0$) that is stronger in males than in females ($|s^m| > |s^f|$).

@@

One might think that when the sex of a zygote is under the control of its mother, there would be strong selection to balance the sex ratio among zygotes. 
However, we find that, as with offspring control, assuming weak selection the invasion fitness of a sex-determiner that is maternally controlled can be written
\begin{equation}
\lambda_{k,XY} \approx 1 + V_A S_A C_k,
\end{equation}
where
%\begin{equation}
%V_A = \frac{- \left[ t_m + h (s_{f}+s_{m}) \right] \left[ t_m + (1-h) (s_{f}+s_{m}) \right]}{(1-2h)^2(s_{f}+s_{m})^2}
%\end{equation}
%is the variance in $A$ ($V_A>0$ when the polymorphic resident equilibrium is stable), 
%with $h$ the dominance coefficient (equal in both sexes), 
$C_k$ is a term that depends on $k$.
%
%%k=0
%\begin{equation}
%C_0 = big \; ugly \; term
%\end{equation}
%
%%k=1
%\begin{equation}
%C_1 = \frac{\left[ (2r(1-R)-R) s^f + (1-2r)R s^m \right]}{2}
%\end{equation}
%
%%k=1/2
%\begin{equation}
%C_{1/2} = even \; bigger, \; uglier \; term
%\end{equation}
%
%When $k=1$ the maternal- and offspring-controlled results are the same, $C_1 = C_W$, since the mutation remains in females (and hence mutant daughters always have mutant mothers).
%However, with $k\neq1$ the results with maternal control differ from those with offspring control.
Of particular interest is $k=1/2$ (i.e., when the mother perfectly balances the sex ratio of her offspring).
When both recombination rates are small we have $C_{1/2} \approx R(s^m - s^f)/8 = \lim_{r \rightarrow 0} C_W/4$.
This implies that, at least under tight linkage, the invasion of maternal ESD is independent of $R$ (because $S_A\propto R^{-1}$) and can invade whenever a neo-$W$ can.

>>>>

%life-cycle
The life-cycle begins with competition between haploid gametes/gametophytes from each sex, where selection depends on the sex of the diploid they came from and their allele at the \textbf{A} locus.
Gametes/gametophytes from males then pair randomly with gametes/gametophytes from females.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (or the \textbf{M} genotype of their mother in the case of maternal control).
Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
This is followed by meiosis with recombination. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$. %is chi really the prob of rec in our recursion equations?
Any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$.
We track the frequency of haploid genotypes produced by each sex from one generation to the next (recursion equations in Sup.\ Mat.).

@@

\subsection*{Resident equilibrium and stability}
With allele $M$ fixed, we follow the dynamics of the frequency of $A$ in gametes/gametophytes from the homogametic sex (e.g., in eggs from an XX female, $p^f_X$) and the frequencies of $A$ in the two types of gametes/gametophytes from the heterogametic sex (e.g., in sperm from an XY male that are X-bearing, $p^m_X$, and Y-bearing, $p^m_Y$).
Assuming selection is weak relative to recombination, the differences in the frequencies of $A$ in each type of gamete/gametophyte are small, and we can solve for the mean frequency of $A$ across all types ($p_A$) as well as the difference in the frequencies of $A$ between two of the three types, to first order in selection.
Linear stability analysis can then be used to determine the stability of this equilibrium.
With no haploid selection our results reduce to those of \cite{vanDoorn:2007eu}. %when $k=0$ (neo-$Y$ invading a $ZW$ system) or when $k=1$ (neo-$W$ invading an $XY$ system) (\textcolor{red}{CHECK}).
However, with haploid selection a stable polymorphism at locus \textbf{A} no longer requires sexually antagonistic selection, as it can also be achieved by ploidally-antagonistic selection.

\subsection*{Sex chromosome turnover}
The spread of a rare mutant $m$ at the \textbf{M} locus in such a population is determined by the leading eigenvalue, $\lambda$, of the system described by the equations for the next generation frequency of haploid genotypes with the mutation. %, evaluated at the above equilibrium.
Below we present the results for invasion into an $XY$ system assuming no competition among gametes/gametophytes from females and linear arrangement \textbf{X}\textbf{A}\textbf{M}. %(i.e., $\chi = r R$). %(SAY SOMETHING HERE ABOUT OTHER ARRANGEMENTS?). %equal dominance coefficients in the two sexes,
Equivalent results for invasions into a $ZW$ system can be derived by consistently switching the roles of males and females \citep{vanDoorn:2010hu}.

@@

Otherwise, the new sex-determining allele increases in frequency on one \textbf{A} background and declines on the other, and invasion requires
\begin{equation}\label{eq:neoYR}
R \left[ \frac{p^f_X w^m_a}{g_A^*} + \frac{(1 - p^f_X) w^m_A}{g_a^*} \right] w^m_{Aa} < \nu^m,
\end{equation}
for the neo-$Y$, and 
\begin{equation}\label{eq:neoWR}
R \left[  \frac{(1 - \bar{p}^m) w^m_{a}}{g_A^*} + \frac{\bar{p}^m w^m_{A}}{g_a^*} \right] w^f_{Aa} < \nu^f
\end{equation}
for the neo-$W$.
Here $\bar{p}^m = (p^m_X+p^m_Y)/2$ is the average frequency of $A$ in gametophytes produced by males, $w^i_{j}$ is the relative viability fitness of sex $i\in\{m,f\}$ depending on their haploid or diploid genotype at the \textbf{A} locus (with $j\in\{A,a,AA,Aa,aa\}$) and
$\nu^i = 
  p^f_X p^m_{\textbf{X}^i} w^m_{a} w^i_{AA} +
  p^f_X (1-p^m_{\textbf{X}^i}) w^m_{a} w^i_{Aa} +
  (1-p^f_X) p^m_{\textbf{X}^i} w^m_{a} w^i_{Aa} +
  (1-p^f_X) (1-p^m_{\textbf{X}^i}) w^m_a w^i_{aa}$,
with $\textbf{X}^m = Y$ and $\textbf{X}^f = X$, is the mean relative fitness of resident individuals of sex $i$.
%and $g_j^*$ is the growth rate of $X^m$ and $Ym$ haplotypes when on an $A$ ($j=A$)  or $a$ background ($j=a$).

The new sex-determining allele is therefore expected to invade for any recombination rate, $R$, when the net flow of double recombinants is from the less fit to the more fit \textbf{A} background (making the terms inside the square brackets in Equations \ref{eq:neoYR} and \ref{eq:neoWR} negative).
%Assuming $g_A^* > 0$ and $g_a^* < 0$ without loss of generality, this is $-(1 - p^f_X) w_{m,A} g_a^* < p^f_X w_{m,a} g_A^*$.
When the net flow of double recombinants is from the more fit to the less fit haplotype, the new sex-determining allele can still invade when the rate of recombination between it and the selected locus, $R$, is small enough.

%weak selection
Assuming weak selection, we can solve for the invasion fitness of the new sex-determining allele, $m$, into the ancestral $XY$ system explicitly, giving
\begin{equation}
\lambda_{m,XY} \approx 1 + V_A S_A C_m,
\end{equation}
where $V_A = p_A(1-p_A)$ is the variance at the \textbf{A} locus, $S_A = s^f (t^m)^2 / \left[ r R \left( s^f + s^m \right)^2 \right]$ (\textcolor{red}{INTERPRETATION?}), $s^f$ and $s^m$ are the respective selection coefficients for $A$ in diploid females and males, $t^m$ is the selection coefficient for $A$ in gametes/gametophytes from males, and we've assumed equal dominance coefficients in the two sexes.

For the neo-$Y$ we have $C_Y = (r - R)s^f $.
The neo-$Y$ can therefore invade whenever it is in tighter linkage with the selected locus than the ancestral sex-determining locus, $r>R$, provided locus \textbf{A} is polymorphic ($V_A>0$) and there is selection among both females ($s^f\neq0$) and gametes/gametophytes from males ($t^m\neq0$).
This differs from \cite{vanDoorn:2007eu}

For the neo-$W$ we have $C_W = [(2r(1-R)-R) s^f + (1-2r)R s^m]/2$.
In this case, even when the novel sex-determining locus is in looser linkage with the selected locus than the ancestral sex-determining locus is, $r<R$, a neo-$W$ can invade.
For example, with $R=1/2$ the neo-$W$ invades if there is any linkage between the ancestral sex-determining and selected loci ($r<1/2$), there is selection among gametes/gametophytes in males ($t^m\neq0$), and there is selection for or against $A$ in both males and females ($s^m s^f >0$) that is stronger in males than in females ($|s^m| > |s^f|$).

@@

One might think that when the sex of zygotes is under the control of mothers, there would be strong selection to balance the sex ratio among zygotes. 
However, we find that, as with offspring control, assuming weak selection the invasion fitness of a sex-determiner that is maternally controlled can be written
\begin{equation}
\lambda_{k,XY} \approx 1 + V_A S_A C_k,
\end{equation}
where
%\begin{equation}
%V_A = \frac{- \left[ t_m + h (s_{f}+s_{m}) \right] \left[ t_m + (1-h) (s_{f}+s_{m}) \right]}{(1-2h)^2(s_{f}+s_{m})^2}
%\end{equation}
%is the variance in $A$ ($V_A>0$ when the polymorphic resident equilibrium is stable), 
%with $h$ the dominance coefficient (equal in both sexes), 
$C_k$ is a term that depends on $k$.
%
%%k=0
%\begin{equation}
%C_0 = big \; ugly \; term
%\end{equation}
%
%%k=1
%\begin{equation}
%C_1 = \frac{\left[ (2r(1-R)-R) s^f + (1-2r)R s^m \right]}{2}
%\end{equation}
%
%%k=1/2
%\begin{equation}
%C_{1/2} = even \; bigger, \; uglier \; term
%\end{equation}
%
%When $k=1$ the maternal- and offspring-controlled results are the same, $C_1 = C_W$, since the mutation remains in females (and hence mutant daughters always have mutant mothers).
%However, with $k\neq1$ the results with maternal control differ from those with offspring control.
Of particular interest is $k=1/2$ (i.e., when the mother perfectly balances the sex ratio of her offspring).
When both recombination rates are small we have $C_{1/2} \approx R(s^m - s^f)/8 = \lim_{r \rightarrow 0} C_W/4$.
This implies that, at least under tight linkage, the invasion of maternal ESD is independent of $R$ (because $S_A\propto R^{-1}$) and can invade whenever a neo-$W$ can.
